---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# waterquality

The goal of **waterquality** is to convert satellite-reflectance imagery to a host of pre-defined water quality algorithms designed for the detection of cholorophyll-a, blue-green algae (Phycocyanin), and Turbidity. 
This package is able to process the following sensor configurations: WorldView-2, Sentinel-2, Landsat-8, MODIS, and MERIS sensors.

## Installation

You can install waterquality from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("RAJohansen/waterquality")
```

## Example

The main function in this package is `wq_calc()`:

```{r example}
library(waterquality)
library(raster)
s2 = stack(system.file("raster/S2_Taylorsville.tif", package = "waterquality"))
s2_two_alg = wq_calc(s2, alg = c("TurbChip09NIROverGreen", "Am092Bsub"), sat = "sentinel2")
plot(s2_two_alg)
```

## Contribution

Feel free to submit issues and enhancement requests.
